A Standardized Effect Size for Evaluating the Strength of Phylogenetic Signal, and Why Lambda is not Appropriate

### 2020-05-11 Manuscript/Manuscript-6-EKB.tex
EKB's edits on Manuscript 6 (based on manuscript-6-dca.rmd)

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers

\author{}
\date{\vspace{-2.5em}}

@@

\hypertarget{a-standardized-effect-size-for-evaluating-the-strength-of-phylogenetic-signal-and-why-lambda-is-not-appropriate}{%
\section{A Standardized Effect Size for Evaluating the Strength of
Phylogenetic Signal, and Why Lambda is not
Appropriate}\label{a-standardized-effect-size-for-evaluating-the-strength-of-phylogenetic-signal-and-why-lambda-is-not-appropriate}}

@@

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. In this study, we evaluated the
efficacy of one commonly used parameter (Pagel's \(\lambda\)) to
estimate the strength of phylogenetic signal in phenotypic traits, and
evaluate the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal. We find that the precision of \(\lambda\)
in estimating actual levels of phylogenetic signal is often inaccurate,
and that biological interpretations of the strength of phylogenetic
signal based on \(\lambda\) are therefore compromised. We then propose a
standardized effect size based on \(\kappa\) (\(Z_\kappa\)), which
measures the strength of phylogenetic signal, and places it on a common
scale for statistical comparison. Tests based on \(Z_\kappa\) provide a
mechanism for formally comparing the strength of phylogenetic signal
across datasets, in much the same manner as effect sizes may be used to
summarize patterns in quantitative meta-analysis. Our approach extends
the phylogenetic comparative toolkit to address hypotheses that compare
the strength of phylogenetic signal between various phenotypic traits,
even when those traits are found in different evolutionary lineages or
have different units or scale.

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets, including measures of serial independence
(\(\mathbf{C}\): Abouheif 1999), autocorrelation estimates (\(I\):
Gittleman and Kot 1990), statistical ratios of trait variation relative
to what is expected given the phylogeny (\(\kappa\): Blomberg et al.
2003; Adams 2014a), and scaling parameters used in maximum likelihood
fitting of the data to the phylogeny (\(\lambda\): Pagel 1999), among
others (e.g., Klingenberg and Gidaszewski 2010). The statistical
properties of these methods -- namely type I error rates and power --
have also been investigated to determine when phylogenetic signal can be
detected and under what conditions (e.g., Münkemüller et al. 2012;
Pavoine and Ricotta 2012; Diniz-Filho et al. 2012; Adams 2014a;
Molina-Venegas and Rodríguez 2017; see also Revell et al. 2008; Revell
2010). One of the most widely used methods for characterizing
phylogenetic signal in macroevolutionary studies is Pagel's \(\lambda\)
(Pagel 1999). The (\(\lambda\)) parameter transforms the lengths of the
internal branches of the phylogeny to improve a fit of data to the
phylogeny via maximum likelihood (Pagel 1999; Freckleton et al. 2002).
Pagel's \(\lambda\) ranges from \(0\to1\), with larger values signifying
a greater dependence of observed trait variation on the phylogeny.
Pagel's \(\lambda\) also has the appeal that it may be included in
phylogenetic generalized least-squares regression (PGLS) to account for
the degree of phylogenetic signal in comparative analyses (see
Freckleton et al. 2002). \hfill\break

In addition to functioning as a parameter that is tuned for appropriate
analysis, \(\lambda\) can function as a descriptive statistic to
describe the relative strength of phylogenetic signal in phenotypic
traits, to determine the extent to which shared evolutionary history has
influenced trait covariation among taxa. The appeal of \(\lambda\) as a
descriptive statistic for evolutionary biologists is a basis for
interpreting ``weak'' versus ``strong'' phylogenetic signal; i.e., small
versus large values of \(\lambda\), respectively, in a comparative sense
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Indeed, statements regarding the strength of phylogenetic signal based
on \(\lambda\) are rather common in the evolutionary literature. For
instance, of the 204 papers published in 2019 that estimated and
reported Pagel's \(\lambda\) (found from a literature survey we
conducted in Google.scholar), 40\% interpreted the strength of
phylogenetic signal for at least one phenotypic trait. Further, because
nearly half of the 1,572 \(\lambda\) values reported were near 0 or 1
(Figure 1) where the biological interpretation of \(\lambda\) is known,
this percentage is even higher. \hfill\break

@@

Various other approaches use \(\lambda\) as a parameter that can be
varied for inferences akin to sensitivity analysis. For instance, some
have performed likelihood ratio tests that compare observed model fits
to those obtained when \(\lambda=0\) or \(\lambda=1\) (Freckleton et al.
2002; Cooper et al. 2010; Bose et al. 2019) or evaluated whether
observed \(\lambda\) differs from an expected \(\lambda\), based on
confidence intervals generated for the expected value (Vandelook et al.
2019). Qualitative comparisons of \(\lambda\) estimates have also been
performed for multiple traits on the same phylogenetic tree to infer
whether the strength of phylogenetic signal is greater in one trait as
compared to another (e.g., Liu et al. 2019; Bai et al. 2019).
\hfill\break

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also implies that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we lack a
general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure of the magnitude of a statistical effect in data, represented on
a common scale (Glass 1976; Cohen 1988). Effect sizes have widespread
use in many areas of the quantiative sciences, as they represent
measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunatley, not all model parameters and descriptive statistics are
effect sizes, and thus many summary measures must first be converted to
statistics with standardized units (i.e., conversion to an effect size)
for meaningful comparison (see Rosenthal 1994). As a consequence, it
follows that only if \(\lambda\) is a statistical effect size can
comparisons of estimates across datasets be interpretable. For the case
of \(\lambda\), this has not yet been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable indicator of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets, and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(\kappa\). Through simulations we
find that the precision of effect sizes based on \(\lambda\)
(\(Z_{\lambda}\)) are less reliable than that those based on \(\kappa\)
(\(Z_\kappa\)), implying that \(Z_\kappa\) is a more robust effect size
measure. We also propose a two-sample test statistic that may be used to
compare the strength of phylogenetic signal among datasets, and provide
an empirical example to demonstrate its use. We conclude that estimates
of phylogenetic signal using Pagel's \(\lambda\) are often inaccurate,
and thus interpreting strength of phylogenetic signal in phenotypic
datasets based on this measure is compromised. By contrast, effect sizes
obtained from \(\kappa\) hold promise for characterizing phylogenetic
signal, and for comparing the strength of phylogenetic signal across
datasets.

\hypertarget{methods-and-results}{%
\section{Methods and Results}\label{methods-and-results}}

\hypertarget{the-precision-of-lambda-is-variable}{%
\subsection{\texorpdfstring{\emph{The Precision of \(\lambda\) is
Variable}}{The Precision of \textbackslash lambda is Variable}}\label{the-precision-of-lambda-is-variable}}

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
published in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these mid-range empirical estimates were explicitly interpreted
in terms of the strength of phylogenetic signal that they represented
(i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression and ANOVA, the results of our simulations were quite
similar. Regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3; see also Supporting Information), and the
spread of \(\lambda_{est}\) was similar to that observed when
\(\lambda\) was estimated for only the dependent variable, as in Figure
2. Taken together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize observed levels of phylogenetic signal in
phenotypic datasets, and that biological interpretations of the strength
of phylogenetic signal based on \(\lambda\) may be highly inaccurate.
\hfill\break

@@

\hypertarget{a-standardized-effect-size-for-phylogenetic-signal}{%
\subsection{\texorpdfstring{\emph{A Standardized Effect Size for
Phylogenetic
Signal}}{A Standardized Effect Size for Phylogenetic Signal}}\label{a-standardized-effect-size-for-phylogenetic-signal}}

@@

To evaluate the utility of \(Z_\kappa\) and \(Z_\lambda\) we calculated
both effect sizes for the simulated datasets generated above, and
summarized the precision of each using its variance
(\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\), Figure 4:
additional results in the Supporting Information). Here two things are
evident. First, estimates of \(Z_\kappa\) linearly track the input
phylogenetic signal whereas estimates of \(Z_\lambda\) do not (Figure
4A, B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_\kappa\). Second, the precision of \(Z_\kappa\) is considerably more
stable as compared with \(Z_\lambda\). This may be seen by calculating
the coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\)) across input
levels of phylogenetic signal. Coefficients of variation in the
precision of \(Z_\kappa\) were up to an order of magnitude smaller for
than for \(Z_\lambda\) (Figure 4C), implying that estimates of the
strength of phylogenetic signal were more reliable and robust when using
\(Z_\kappa\). \hfill\break

@@

\hypertarget{statistical-comparisons-of-phylogenetic-signal}{%
\subsection{\texorpdfstring{\emph{Statistical Comparisons of
Phylogenetic
Signal}}{Statistical Comparisons of Phylogenetic Signal}}\label{statistical-comparisons-of-phylogenetic-signal}}

@@

\hypertarget{empirical-example}{%
\subsection{\texorpdfstring{\emph{Empirical
Example}}{Empirical Example}}\label{empirical-example}}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\usepackage[normalem]{ulem}
% avoid problems with \sout in headers with hyperref:
\pdfstringdefDisableCommands{\renewcommand{\sout}{}}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers
% https://github.com/rstudio/rmarkdown/issues/337
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

% https://github.com/rstudio/rmarkdown/pull/252
\usepackage{titling}
\setlength{\droptitle}{-2em}

\pretitle{\vspace{\droptitle}\centering\huge}
\posttitle{\par}

\preauthor{\centering\large\emph}
\postauthor{\par}

\predate{\centering\large\emph}
\postdate{\par}

\date{}

@@

\section{A Standardized Effect Size for Evaluating the Strength of
Phylogenetic Signal, and Why Lambda is not
Appropriate}\label{a-standardized-effect-size-for-evaluating-the-strength-of-phylogenetic-signal-and-why-lambda-is-not-appropriate}

@@

\section{Abstract}\label{abstract}

Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. In this study, we evaluated the
efficacy of one commonly used parameter (Pagel's \(\lambda\)) to
estimate the strength of phylogenetic signal in phenotypic traits, and
evaluate the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal. We find that the precision of \(\lambda\)
in estimating actual levels of phylogenetic signal is often inaccurate,
and that biological interpretations of the strength of phylogenetic
signal based on \(\lambda\) are therefore compromised. We then propose a
standardized effect size based on \(\kappa\) (\(Z_\kappa\)), which
measures the strength of phylogenetic signal, and places it on a common
scale for statistical comparison. Tests based on \(Z_\kappa\) provide a
mechanism for formally comparing the strength of phylogenetic signal
across datasets, in much the same manner as effect sizes may be used to
summarize patterns in quantitative meta-analysis. Our approach extends
the phylogenetic comparative toolkit to address hypotheses that compare
the strength of phylogenetic signal between various phenotypic traits,
even when those traits are found in different evolutionary lineages or
have different units or scales.

@@

\section{Introduction}\label{introduction}

@@

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets, including measures of serial independence
(\(\mathbf{C}\): Abouheif 1999), autocorrelation estimates (\(I\):
Gittleman and Kot 1990), statistical ratios of trait variation relative
to what is expected given the phylogeny (\(\kappa\): Blomberg et al.
2003; Adams 2014a), and scaling parameters used in maximum likelihood
fitting of the data to the phylogeny (\(\lambda\): Pagel 1999), among
others (e.g., Klingenberg and Gidaszewski 2010). The statistical
properties of these methods -- namely type I error rates and power --
have also been investigated to determine when phylogenetic signal can be
detected and under what conditions (see also Revell et al. 2008; Revell
2010; e.g., Münkemüller et al. 2012; Pavoine and Ricotta 2012;
Diniz-Filho et al. 2012; Adams 2014a; Molina-Venegas and Rodríguez
2017). One of the most widely used methods for characterizing
phylogenetic signal in macroevolutionary studies is Pagel's \(\lambda\)
(Pagel 1999). The \sout{(\(\lambda\))} parameter (\(\lambda\))
transforms the lengths of the internal branches of the phylogeny to
improve \sout{a} THE fit of data to the phylogeny via maximum likelihood
(Pagel 1999; Freckleton et al. 2002). Pagel's \(\lambda\) ranges from
\(0\to1\), with larger values signifying a greater dependence of
observed trait variation on the phylogeny. Pagel's \(\lambda\) also has
the appeal that it may be included in phylogenetic generalized
least-squares regression (PGLS) to account for the degree of
phylogenetic signal in comparative analyses (see Freckleton et al.
2002). \hfill\break

In addition to functioning as a parameter that is tuned for appropriate
analysis, \(\lambda\) can function as a descriptive statistic \sout{to
describe} CHARACTERIZING the relative strength of phylogenetic signal in
phenotypic traits to determine the extent to which shared evolutionary
history has influenced trait covariation among taxa. \sout{The appeal}
THIS DESCRIPTIVE UTILITY of \(\lambda\) \sout{as a descriptive statistic
for evolutionary biologists} is \sout{a} THE basis for interpreting
``weak'' versus ``strong'' phylogenetic signal\sout{,} (i.e., small
versus large values of \(\lambda\), respectively)\sout{,} in a
comparative sense (e.g., De Meester et al. 2019; Pintanel et al. 2019;
Su et al. 2019). Indeed, statements regarding the strength of
phylogenetic signal based on \(\lambda\) are rather common in the
evolutionary literature. For instance, of the 204 papers published in
2019 that estimated and reported Pagel's \(\lambda\) (found from a
literature survey we conducted in Google.scholar), 40\% EXPLICITLY
interpreted the strength of phylogenetic signal for at least one
phenotypic trait. Further, because nearly half of the 1,572 \(\lambda\)
values reported were near 0 or 1 (Figure 1) where the biological
interpretation of \(\lambda\) is known, this percentage is even higher.
\hfill\break

@@

Various other approaches use \(\lambda\) as a parameter that can be
\sout{varied} ADJUSTED for inferences akin to sensitivity analysis. For
instance, some have performed likelihood ratio tests that compare
observed model fits to those obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019) or evaluated whether observed \(\lambda\) differs from an expected
\(\lambda\), based on confidence intervals generated for the expected
value (Vandelook et al. 2019). Qualitative comparisons of \(\lambda\)
estimates have also been performed for multiple traits on the same
phylogenetic tree to infer whether the strength of phylogenetic signal
is greater in one trait as compared to another (e.g., Liu et al. 2019;
Bai et al. 2019). \hfill\break

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., \sout{the}
LOW number of species) was the underlying cause of the increased
variation across parameter estimates (Boettiger et al. 2012). Indeed,
such a pattern is common with statistical estimators, as summary
statistics and parameters are often more precise at greater sample sizes
(Cohen 1988). However, this conclusion also implies that the precision
of \(\lambda\) remains constant across its range
(\(\lambda = 0 \to 1\)); an assumption that to date, has not been
verified. Thus, despite widespread use of Pagel's (1999) \(\lambda\) in
macroevolutionary studies, at present, we lack a general understanding
of the precision with which \(\lambda\) can estimate levels of
phylogenetic signal in phenotypic datasets. \hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure of the magnitude of a statistical effect in data, represented on
a common scale (Glass 1976; Cohen 1988). Effect sizes have widespread
use in many areas of the quantiative sciences, as they represent
measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunatley, not all model parameters and descriptive statistics are
effect sizes, and thus many summary measures must first be converted to
statistics with standardized units (i.e., conversion to an effect size)
for meaningful comparison (see Rosenthal 1994). As a consequence, it
follows that only if \(\lambda\) is a statistical effect size can
comparisons of estimates across datasets be interpretable. \sout{For the
case of \(\lambda\), this} HOWEVER, THE CALCULATION AND STATISTICAL
BEHAVIOR OF \(\lambda\) AS AN EFFECT SIZE has not yet been explored.
\hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable indicator of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(\kappa\). Through simulations we
find that the precision of effect sizes based on \(\lambda\)
(\(Z_{\lambda}\)) are less reliable than that those based on \(\kappa\)
(\(Z_\kappa\)), implying that \(Z_\kappa\) is a more robust effect size
measure. We also propose a two-sample test statistic that may be used to
compare the strength of phylogenetic signal among datasets, and provide
an empirical example to demonstrate its use. We conclude that estimates
of phylogenetic signal using Pagel's \(\lambda\) are often inaccurate,
and thus interpreting strength of phylogenetic signal in phenotypic
datasets based on this measure is compromised. By contrast, effect sizes
obtained from \(\kappa\) hold promise for characterizing phylogenetic
signal, and for comparing the strength of phylogenetic signal across
datasets.

\section{Methods and Results}\label{methods-and-results}

\subsection{\texorpdfstring{\emph{The Precision of \(\lambda\) is
Variable}}{The Precision of \textbackslash{}lambda is Variable}}\label{the-precision-of-lambda-is-variable}

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al. (2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
published in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these mid-range empirical estimates were explicitly interpreted
in terms of the strength of phylogenetic signal that they represented
(i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression and ANOVA, the results of our simulations were quite
similar. Regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3; see also Supporting Information), and the
spread of \(\lambda_{est}\) was similar to that observed when
\(\lambda\) was estimated for only the dependent variable, as in Figure
2. Taken together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize \sout{observed} KNOWN levels of phylogenetic
signal in phenotypic datasets, and that biological interpretations of
the strength of phylogenetic signal based on \(\lambda\) may be highly
inaccurate. \hfill\break

@@

\subsection{\texorpdfstring{\emph{A Standardized Effect Size for
Phylogenetic
Signal}}{A Standardized Effect Size for Phylogenetic Signal}}\label{a-standardized-effect-size-for-phylogenetic-signal}

@@

To evaluate the utility of \(Z_\kappa\) and \(Z_\lambda\) we calculated
both effect sizes for the simulated datasets generated above, and
summarized the precision of each using its variance
(\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\), Figure 4:
additional results in the Supporting Information). Here two things are
evident. First, estimates of \(Z_\kappa\) linearly track the input
phylogenetic signal whereas estimates of \(Z_\lambda\) do not (Figure
4A, B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_\kappa\). Second, the precision of \(Z_\kappa\) is considerably more
stable as compared with \(Z_\lambda\). This may be seen by calculating
the coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\)) across input
levels of phylogenetic signal. Coefficients of variation in the
precision of \(Z_\kappa\) were up to an order of magnitude smaller
\sout{for} than for \(Z_\lambda\) (Figure 4C), implying that estimates
of the strength of phylogenetic signal were more reliable and robust
when using \(Z_\kappa\). \hfill\break

@@

\subsection{\texorpdfstring{\emph{Statistical Comparisons of
Phylogenetic
Signal}}{Statistical Comparisons of Phylogenetic Signal}}\label{statistical-comparisons-of-phylogenetic-signal}

@@

\subsection{\texorpdfstring{\emph{Empirical
Example}}{Empirical Example}}\label{empirical-example}
